---
title: "Get public transportation data"
output: html_notebook
---

Get all nessecary data for public transportation:
 * all stops in Tallinn
 * identify tram stops
 * do API call for tram arrival info in stops

```{r}
library(sf)
library(tidyverse)
library(httr)
library(rvest)
library(seleniumPipes)
library(osmdata)
```

## Public transportation stops data

Download public transportation stops in Tallinn with stop id (for API calls)
As the data is in some weird java script format some manual formatting is necessary (add quotes to all keywords: type, coordinates etc)
Source data is form a website: https://gis.ee/tallinn/
```{r}
download.file("https://gis.ee/tallinn/peatused.js?ver=2017-09-18T2314", "data/peatused_raw.txt")
```

Read public transportation stops data as a sf object and plot it
```{r}
peatused <- st_read("data/peatused.txt")

mapview::mapview(peatused)
```


## API call for every stop

Generate url for every stop to ma API calls
```{r}
all_stop_id_url <- peatused %>% 
  as.tibble() %>% 
  select(id = SiriID) %>% 
  distinct(id) %>% 
  mutate(api_url = str_c("https://transport.tallinn.ee/siri-stop-departures.php?stopid=", id))
```

Function to download API call information for a given stop url
```{r}
read_csv_char <- function(x){
  raw <- read_csv(x, col_types = "cccccc",
                  col_names = c("transport", "route_num", "expected_time_in_seconds",
                  "schedule_time_in_seconds", "stop_name", "empty"))
  raw %>%
    select(-empty) %>%
    filter(!transport %in% c("stop", "Transport")) %>% 
    mutate(url = x)
}

# ignore errors
read_csv_char_possibly <- possibly(read_csv_char, NULL)
```

Make the API call for all public transportation stops.
It is necessary to identify all tram stops.
```{r, warning=FALSE, message=FALSE}
all_stops_raw <- map_df(all_stop_id_url$api_url, read_csv_char_possibly)
```

Filter only tram stops
```{r}
tram_stops <- all_stops_raw %>%
  filter(transport == "tram") %>% 
  distinct(url, stop_name) %>% 
  left_join(all_stop_id_url, by = c("url" = "api_url"))
```

Tram stops as a sf object with coordinates and visualized on a map
```{r}
tram_stops_sf <- peatused %>% 
  inner_join(tram_stops, by = c("SiriID" = "id")) 

tram_stops_sf %>% 
  mapview::mapview()
```

Add calculated fields to tram arrival data.
Also test data with realtime gps data in https://gis.ee/tallinn/
```{r}
all_stops_raw %>%
  filter(transport == "tram") %>% 
  mutate(late = as.numeric(expected_time_in_seconds) - as.numeric(schedule_time_in_seconds),
         expected_hms = hms::as.hms(as.numeric(expected_time_in_seconds)),
         schedule_hms = hms::as.hms(as.numeric(schedule_time_in_seconds))) %>% 
  left_join(all_stop_id_url, by = c("url" = "api_url")) %>% 
  left_join(peatused %>% as.tibble(), by = c("id" = "SiriID")) %>% 
  filter(Name == "L. Koidula",
         route_num == 1)

  # filter(late == -629)
  # count(late, sort = TRUE) %>% 
  # arrange(late)
  ggplot(aes(late)) +
  geom_histogram(binwidth = 10) +
  coord_cartesian(xlim = c(-500, 500))

```

Save tram stops 
```{r}
tram_stops %>% 
  select(url) %>% 
  saveRDS("data/tram_stops.rds")
```

Save additional data
```{r}
save(tram_stops, tram_stops_sf, peatused, all_stop_id_url, all_stops_raw,
     file = "data/arrival_data.RData")
```


Test snapshot call for all tram stops
```{r, warning=FALSE, message=FALSE}
system.time(
 tram_test_1 <- map_df(tram_stops_sf$url, read_csv_char_possibly) 
)
```

```{r}
t <- readRDS("data/arrival/1523873762.rds")

load("data/arrival_data.RData")
```


```{r}
t %>%
  mutate(late = as.numeric(expected_time_in_seconds) - as.numeric(schedule_time_in_seconds),
         expected_hms = hms::as.hms(as.numeric(expected_time_in_seconds)),
         schedule_hms = hms::as.hms(as.numeric(schedule_time_in_seconds))) %>% 
  left_join(all_stop_id_url, by = c("url" = "api_url")) %>% 
  left_join(peatused %>% as.tibble(), by = c("id" = "SiriID")) %>% 
  arrange(desc(late))
  filter(Name == "L. Koidula",
         route_num == 1)

  # filter(late == -629)
  # count(late, sort = TRUE) %>% 
  # arrange(late)
  ggplot(aes(late)) +
  geom_histogram(binwidth = 10) +
  coord_cartesian(xlim = c(-500, 500))
```

## Query tram timetable data

Url for every tram line timetable
```{r}
timetable_url <- c(
  "https://transport.tallinn.ee/#tram/1/a-b",
  "https://transport.tallinn.ee/#tram/1/b-a",
  "https://transport.tallinn.ee/#tram/2/a-b",
  "https://transport.tallinn.ee/#tram/2/b-a",
  "https://transport.tallinn.ee/#tram/3/a-b",
  "https://transport.tallinn.ee/#tram/3/b-a",
  "https://transport.tallinn.ee/#tram/4/a-b",
  "https://transport.tallinn.ee/#tram/4/b-a"
)
```


Esmalt tuleb Ubuntu puhul käivitada Seleniumi server läbi shelli kahe käsurea:

$ sudo docker run -d -p 4445:4444 selenium/standalone-firefox
$ sudo docker ps

Seleniumi serveri sulgemiseks (käsurea lõpus olev nr on Container ID, mille saab käivituse käsust $ sudo docker ps):
$ sudo docker stop 7e8466a91673

Seejärel saab ühenduse üles ning alustada veebilehe kraapimist.

Create Selenium connection
```{r}
remDr <- remoteDr(port = 4445L)
```

Stop Selenium connection
```{r}
# remDr %>%
#   deleteSession()
```

Function to get page source for each tram line to extract href to timetable
```{r}
get_timetable_href <- function(x){
  # pause for 0.5-1.5 sec
  Sys.sleep(runif(1, 0.5, 1.5))
  
  page_timetable <- remDr %>% 
    go(x) %>% 
    getPageSource()
  
  page_timetable %>% 
    html_nodes(".days-12345 a") %>%
    html_attr("href") %>% 
    str_c("https://transport.tallinn.ee/", .) %>% 
    as.tibble() %>% 
    rename(url = value)
}

# ignore errors
get_timetable_href_possibly <- possibly(get_timetable_href, NULL)
```

Query all tram lines timetable url
```{r}
tram_line_timetable_url <- map_df(timetable_url, get_timetable_href_possibly)
```

Function to query info for every tramline and every stop timetable
```{r}
get_tram_timetable <- function(x){
  # pause for 0.5-1.5 sec
  Sys.sleep(runif(1, 0.5, 1.5))
  
  timetable_raw <- remDr %>% 
    go(x) %>% 
    getPageSource() %>% 
    html_node("#dlDirStops1") %>%
    html_nodes(".hover") %>% 
    html_text()
  
  timetable <- tibble(time_stop = timetable_raw, url = x)
  
  return(timetable)
}

get_tram_timetable_possibly <- possibly(get_tram_timetable, NULL)
```

Get timetables for every tramline and every stop
```{r}
system.time(
  tram_timetable_raw <- map_df(tram_line_timetable_url$url, get_tram_timetable_possibly)
)
```

Extract time, stop and tram line info from raw data
```{r}
tram_timetable_processed <- tram_timetable_raw %>% 
  mutate(time = str_extract(time_stop, "[[:digit:]]{1,2}:[[:digit:]]{1,2}"),
         stop = str_squish(str_replace_all(time_stop, time, "")),
         route_num = str_extract(url, "(?<=#tram/)[[:digit:]]"))

tram_timetable_processed
```

Add last stop to every tram route
```{r}
tram_direction <- tram_timetable_processed %>% 
  group_by(url) %>% 
  filter(row_number() == max(row_number())) %>% 
  select(url, last_stop = stop)

tram_timetable <- tram_timetable_processed %>% 
  left_join(tram_direction, by = "url") %>% 
  select(route_num, stop, last_stop, time, url)
```

Save tram timetable
```{r}
save(tram_timetable, file = "data/tram_timetable.rData")
```

## Query tram lines sf object

Query tram lines from OSM
```{r}
# Estonian projection
eesti_proj4 <- "+proj=lcc +lat_1=59.33333333333334 +lat_2=58 +lat_0=57.51755393055556 +lon_0=24 +x_0=500000 +y_0=6375000 +ellps=GRS80 +towgs84=0,0,0,0,0,0,0 +units=m +no_defs"

tram_lines <- getbb("tallinn") %>%
  opq() %>%
  add_osm_feature("railway", value = "tram") %>% 
  osmdata_sf() %>%
  .$osm_lines %>% 
  st_transform(., crs = eesti_proj4)

mapview::mapview(tram_lines)
```

Save tram lines sf object
```{r}
save(tram_lines, file = "data/tram_lines_sf.rData")
```

