---
title: "Prepare tram arrival data for analysis"
output: html_notebook
---

```{r setup}
library(tidyverse)
library(lubridate)
library(hms)
library(sf)
library(mapview)
library(mapedit)
library(leaflet)

load("data/arrival_data.RData")
load("data/tram_timetable.rData")
load("data/tram_lines_sf.rData")
```

Load whole arrival data from API calls
```{r}
# fail path for all arrival data
files = list.files("data/arrival", full.names = TRUE)

# load arrival data into one dataframe
system.time(
  tram_arrival_raw <- map_df(files, readRDS)
)
```

inspect arrival raw data
```{r}
glimpse(tram_arrival_raw)
```

Prepare arrival data so, that for every stop and tram there is one row (last) in table
```{r}
tram_arrival <- tram_arrival_raw %>% 
  # remove duplicated rows
  arrange(route_num, url, schedule_time_in_seconds, expected_time_in_seconds,
          desc(timestamp)) %>%
  distinct(route_num, url, schedule_time_in_seconds, expected_time_in_seconds,
           .keep_all = TRUE) %>%
  # concert to proper timestamp of API call
  mutate(timestamp = as.POSIXct(timestamp, origin = "1970-01-01"),
         scheduled_vs_expected = as.numeric(expected_time_in_seconds) - as.numeric(schedule_time_in_seconds),
         expected_hms = as.hms(as.numeric(expected_time_in_seconds)),
         schedule_hms = as.hms(as.numeric(schedule_time_in_seconds)),
         # round down to minute
         schedule_hms_round = as.hms(str_replace(schedule_hms, "[[:digit:]]{2}$", "00"))) %>% 
  # add stop id
  left_join(all_stop_id_url, by = c("url" = "api_url")) %>% 
  # add stop name
  left_join(peatused %>% as.tibble() %>% select(-geometry), by = c("id" = "SiriID")) %>% 
  # arrival data for one 24H period
  # remove trips to depoo
  filter(timestamp >= dmy_hm("16.04.2018 01:00"),
         timestamp <= dmy_hm("17.04.2018 01:00"),
         !str_detect(stop_name, "depoo")) %>%
  # for every stop and tram keep only last delay data
  arrange(id, route_num, stop_name, schedule_time_in_seconds, desc(timestamp)) %>% 
  distinct(id, route_num, stop_name, schedule_hms_round,
           .keep_all = TRUE) %>% 
  rename(last_stop = stop_name, stop = Name) %>% 
  # format to proper time hh:mm
  mutate(stop = as.character(stop)) %>% 
  select(-url)

glimpse(tram_arrival)
```

```{r}
tram_timetable %>% 
  glimpse()
```

Combine tram timetable data with actual arrival data
```{r}
tram_timetable_arrival <- tram_timetable %>% 
  # round down to nearest minute to join datasets
  mutate(schedule_hms_round = as.hms(str_c(str_c(time, ":00")))) %>% 
  filter(last_stop != "Vineeri") %>% 
  # remove duplicates for scheduled time
  arrange(route_num, stop, last_stop, schedule_hms_round) %>% 
  distinct(route_num, stop, last_stop, time, schedule_hms_round, .keep_all = TRUE) %>% 
  # join arrival data
  left_join(tram_arrival, by = c("route_num", "stop", "last_stop", "schedule_hms_round")) %>% 
  arrange(url, schedule_hms_round) %>% 
  select(-transport, -schedule_hms, -expected_time_in_seconds,
         -schedule_time_in_seconds, -id) %>% 
  # fill missin timestamp and scheduled_vs_expected columns based on last known value
  group_by(url) %>% 
  fill(timestamp, scheduled_vs_expected) %>% 
  group_by(url) %>% 
  fill(timestamp, scheduled_vs_expected, .direction = "up") %>% 
  ungroup() %>% 
  # if expected time is missing then calculate it from scheduled time and 
  # scheduled_vs_expectd difference from last known value
  mutate(expected_hms_temp = as.numeric(schedule_hms_round + hms(scheduled_vs_expected)),
         expected_hms = coalesce(expected_hms, expected_hms_temp)) %>% 
  select(-expected_hms_temp) 

glimpse(tram_timetable_arrival)
```

Remove anomalities:
 * if one timestamp is different from previous and next timestam more than 900sec, then use previous timestamp and corresponding scheduled vs expected arrival time differenece
 * in case of first and last stop compare only with next or previous timestamp and correct with previous or next value
 * if scheduled vs expected arrival time differenece is different more than 300sec from previous and next time difference, then also correct timestamp and time difference value with previous or next actual value.
```{r}
tram_timetable_arrival_fixed <- tram_timetable_arrival %>% 
  group_by(url) %>%
  # add special feature for first and last stop
  mutate(feature = ifelse(row_number() == max(row_number()), "last",
                          ifelse(row_number() == min(row_number()), "first", NA))) %>% 
  group_by(url) %>% 
  # calculate timestamp and sched_vs_expect difference with previous and next value
  # use previously calculated features to decide wether value is anomality 
  mutate(timestamp_diff_lag = timestamp - lag(timestamp),
         timestamp_diff_lead = timestamp - lead(timestamp),
         sched_vs_expect_lag = scheduled_vs_expected - lag(scheduled_vs_expected),
         sched_vs_expect_lead = scheduled_vs_expected - lead(scheduled_vs_expected),
         difference = ifelse(((timestamp_diff_lag > 900 | timestamp_diff_lag < -900) & 
                         (timestamp_diff_lead > 900 | timestamp_diff_lead < -900)) |
                         (feature == "first" & (timestamp_diff_lead > 900 | timestamp_diff_lead < -900)) |
                         (feature == "last" & (timestamp_diff_lag > 900 | timestamp_diff_lag < -900)) |
                         ((sched_vs_expect_lag > 300 | sched_vs_expect_lag < -300) & 
                         (sched_vs_expect_lead > 300 | sched_vs_expect_lead < -300)) |
                         (feature == "first" & (sched_vs_expect_lead > 600 | sched_vs_expect_lead < -600)) |
                         (feature == "last" & (sched_vs_expect_lag > 600 | sched_vs_expect_lag < -600)), 1, NA)) %>%
  ungroup() %>% 
  # remove timestamp, scheduled_vs_expected and expected_hms values for anomal rows
  mutate(timestamp = ifelse(is.na(difference), timestamp, NA),
         scheduled_vs_expected = ifelse(is.na(difference), scheduled_vs_expected, NA),
         expected_hms = ifelse(is.na(difference), expected_hms, NA)) %>% 
  # fill missin timestamp and scheduled_vs_expected columns based on last known value
  group_by(url) %>% 
  fill(timestamp, scheduled_vs_expected) %>% 
  group_by(url) %>% 
  # fill missin timestamp and scheduled_vs_expected columns based on next known value
  fill(timestamp, scheduled_vs_expected, .direction = "up") %>% 
  ungroup() %>% 
  # if expected time is missing then calculate it from scheduled time and 
  # scheduled_vs_expectd difference from last known value
  mutate(expected_hms_temp = as.numeric(schedule_hms_round + hms(scheduled_vs_expected)),
         expected_hms = coalesce(expected_hms, expected_hms_temp)) %>% 
  select(-expected_hms_temp, -timestamp_diff_lag, -timestamp_diff_lead, -sched_vs_expect_lag,
         -sched_vs_expect_lead, -feature, -difference) %>% 
  # format correctly timestamp and expected_hms
  mutate(timestamp = as.POSIXct(timestamp, origin = "1970-01-01"),
         expected_hms = as.hms(expected_hms))

glimpse(tram_timetable_arrival_fixed)
```

Fix column names for tram stops SF dataframe
```{r}
tram_stops_sf_fixed <- tram_stops_sf %>% 
  rename(last_stop = stop_name, 
         stop = Name,
         stop_id = SiriID) %>% 
  mutate(stop = as.character(stop)) %>% 
  select(-url)
```


## Prepare tram lines sf object

Remove depo lines from tram lines sf object
```{r}
# remove manually tram lines in depos
tram_lines_no_depo <- mapview(tram_lines) %>% 
  editMap("tram_lines")

# Estonian projection
eesti_proj4 <- "+proj=lcc +lat_1=59.33333333333334 +lat_2=58 +lat_0=57.51755393055556 +lon_0=24 +x_0=500000 +y_0=6375000 +ellps=GRS80 +towgs84=0,0,0,0,0,0,0 +units=m +no_defs"

# Create a buffer from manually deletd objects
tram_lines_depo_buffer <- tram_lines_depo$deleted %>% 
  st_transform(., crs = eesti_proj4) %>% 
  st_buffer(., dist = 10)

# tram lines objects, that are within depo buffer
tram_lines_in_depo_buffer <- st_contains(tram_lines_depo_buffer, tram_lines_sf %>% st_transform(., crs = eesti_proj4))

# get row numbers of deleted depo lines
depo_row_numbers <- map_df(seq(1, length(tram_lines_in_depo_buffer), by = 1), 
                          function(x){as.tibble(t[[x]])}) %>% 
  distinct(value) %>% 
  pull(value)

# exclude depo lines from sf object
tram_lines_no_depo <- tram_lines_sf %>% 
  filter(!row_number() %in% depo_row_numbers)

mapview(tram_lines_no_depo)
```

Chane projection of tram lines sf objects to the same as tram stops sf object
```{r}
tram_lines_sf <- st_transform(tram_lines_no_depo, crs = sf::st_crs(tram_stops_sf))
```

PLot on leaflet map tram lines with stops
```{r}
leaflet() %>% 
  addProviderTiles("CartoDB.DarkMatterNoLabels", group = "CartoDB") %>% 
  # addCircleMarkers()
  addPolylines(data = tram_lines_sf,
               color = "orange") %>% 
  addCircles(data = e %>% st_transform(., crs= st_crs(tram_lines_sf)),
             color = "red",
             fillOpacity = 1, 
             radius = 50)
```


Test how to split tram lines from each stop
Unsucessfull!
```{r}
library(lwgeom)

e <- tram_stops_sf %>% 
  distinct(Name, .keep_all = TRUE) %>% 
  st_transform(., crs = eesti_proj4)

e_buffer <- st_buffer(e, dist = 20)

u <- tram_lines_sf %>% 
             st_transform(., crs = eesti_proj4) %>% 
             st_combine() %>% 
  st_union()
    st_cast(., "LINESTRING") %>% 

  st_split(., st_sfc(st_point(c(24.67335, 59.45904)))) %>% 
  st_collection_extract(., "LINESTRING") %>% 
  mapview()

st_crosses()           
           
%>% 
  st_split(., tram_lines_sf %>% 
             st_transform(., crs = eesti_proj4) %>% 
             st_combine() %>% 
             st_collection_extract(., "LINESTRING"))

mapview(tram_lines_sf)

sf::st

?st_split

mls_pts <- tram_lines_sf %>%
  st_cast(., "POINT")
  lapply(., function(x) {
    st_sfc(x) %>% 
      st_cast(., 'POINT')})
```


## Save results

Save results for exploratory analysis
```{r}
save(tram_timetable_arrival_fixed, tram_stops_sf_fixed, file = "data/tram_arrival.rData")
```

