---
layout: post
---


```{r setup, message=FALSE, warning=FALSE, echo=FALSE}
# load packages and pre process data
library(tidyverse)
library(hms)
library(lubridate)
library(hrbrthemes)
library(plotly)

load("data/tram_arrival.rData")

# Add first stop to every tram ride 
tram_timetable_arrival <- tram_timetable_arrival_fixed %>% 
  # add date to correctly arrange rides that occure close to 00:00
  mutate(schedule_hms_round_nr = as.numeric(schedule_hms_round),
         # if hours after 00:00, and before 04.00, then date 17.04, else 16.04
         date = if_else(schedule_hms_round_nr < 15000, ymd("2018.04.17"), ymd("2018.04.16"))) %>% 
  select(-schedule_hms_round_nr) %>% 
  arrange(url, date, schedule_hms_round) %>% 
  group_by(url) %>%
  # add special feature for first stop
  mutate(first_stop = ifelse(row_number() == min(row_number()), stop, NA)) %>% 
  fill(first_stop) %>% 
  ungroup() %>% 
  # time difference between scheduled and expected arrival time in minutes
  mutate(scheduled_vs_expected_min = scheduled_vs_expected / 60,
         # add tram id (route_num_first_stop_last_stop)
         tram_id = str_c(route_num, ": ", first_stop, " - ", last_stop))
```

### {.tabset .tabset-fade}

```{r actual_vs_scheduled_prep, message=FALSE, warning=FALSE, echo=FALSE}
# prepare data and write a function to plot every tram ride

# tram rides after 00:00 and before 02:00
# use list to remove them, as they mix up the plot
late_rides_filter <- tram_timetable_arrival %>% 
  filter(schedule_hms_round < as.hms("02:00:00") | expected_hms < as.hms("02:00:00")) %>% 
  distinct(url) %>% 
  pull(url)

# last stop arrival time for every tram ride
tram_arrival_distribution <- tram_timetable_arrival %>% 
  filter(stop == last_stop) 

# if tram arrives to the last stop more than 2 minutes late, then "late", if more then 2 minutes early, then "early", othervise "on time"
# use categories to color lines on chart
late_arrival <- tram_arrival_distribution %>%
  group_by(tram_id) %>%
  mutate(late_early = ifelse(scheduled_vs_expected_min > 2, "late",
                             ifelse(scheduled_vs_expected_min < -2, "early", "on time"))) %>%
  ungroup() %>%
  select(url, late_early)

# prepare data for ploting
arrival_data_chart_raw <- tram_timetable_arrival %>% 
  # add feature "late", "early" or "on time" for every ride
  left_join(late_arrival, by = "url") %>%
  # remove rides after 00:00
  filter(!url %in% late_rides_filter) %>% 
  ungroup() %>%
  # rename some columns for plotting
  rename(stop_name = stop,
         scheduled_arrival = schedule_hms_round,
         actual_arrival = expected_hms)

# Create a function to plot all rides for selected tram line
plot_tram_arrival <- function(tram_nr = "1"){
  
  # only rides for selected tram line
  arrival_data_chart <- arrival_data_chart_raw %>% 
    filter(str_detect(tram_id, tram_nr))
  
  # find stops that cause more than 2 minutes of delay
  delay_reason <- arrival_data_chart %>%
    group_by(url) %>%
    # find delay in minutes that a ride between two stops has caused
    mutate(delay_cum = scheduled_vs_expected_min - lag(scheduled_vs_expected_min),
           # label those with more than 2 min delay or if the delay is at the first stop already more than 2 min
           delay_reason = ifelse(coalesce(delay_cum, scheduled_vs_expected_min) > 2, 1, NA)) %>%
    filter(delay_reason == 1) %>%
    ungroup()
  
  # plot every ride for selected tram number
  arrival_chart <- arrival_data_chart %>%
    ggplot(aes(drlib::reorder_within(stop_name, scheduled_arrival, tram_id),  # reorder stops
               actual_arrival, group = url, color = late_early,
               # popup text
               text = paste("stop name: ", stop_name,
                            '\nscheduled arrival time: ', str_sub(as.character(scheduled_arrival), start = 0, end = 5),
                            "\nactual arrival time: ", str_sub(as.character(actual_arrival), start = 0, end = 5)))) +
    geom_line() +
    # points for exact stops where delays (>2min) happen
    geom_point(data = delay_reason, 
               aes(drlib::reorder_within(stop_name, scheduled_arrival, tram_id),  # reorder stop names
               actual_arrival, group = url), color = "red") +
    # colors for "late", "early" and "on time" rides
    scale_color_manual(values = c("early" = "blue", "late" = "red", "on time" = "grey35")) +
    # format y scale
    scale_y_reverse(breaks = seq(0, 3600 * 24, 3600),
                    labels = function(x) str_c(hour(as.hms(x)), ":", minute(as.hms(x)), "0")) +
    # custom theme
    theme_ipsum_rc() +
    # group by tram_id
    facet_wrap(~ tram_id, ncol = 6, scales = "free_x") +
    theme(panel.grid.major = element_line(colour = "#f0f0f0"),
          axis.text.x = element_blank(),  # no x axis labels
          axis.title.x = element_blank(),  # no x axis title
          legend.position = "top",
          legend.title = element_blank()) +  # no title for legend
    labs(title = str_c("Tram number ", tram_nr, ": every trip on 16.04.2018"))
  
  # generate interactive plotly chart
  ggplotly(arrival_chart, 
           height = 1500, width = 1000,  # chart size
           tooltip = c("text")) %>%  # tooltip text
    # format legend position
    layout(legend = list(orientation = "h", x = 0.35, y = 1.04))
}
```

#### Tram number 1 
<br>
What's really interesting on this chart is a delay that started around 14:30 between Hobujaama and Tallinna Ülikool stop and affected in total 4 trams. The delayed time accumulated again on the same place on the trip back to Kopli. All together the delays affected those trams for about 2 hours and 4 trips. 

On the rush hour there seems to be no special delays. For some reason 5 trams are late to the stop J. Poska in the very beginning of the trip after 18:00. They all make up the delay during the rest of the ride, but for the first half of their trip they are late to every stop.

```{r tram_1, echo=FALSE}
plot_tram_arrival("1")
```

#### Tram number 2 
<br>
Tram number 2 has by far the most rides that arrive too early to stops. Trams arriving too early could be even more frustrating than trams being late a little bit. People who schedule their rides with timetables could easily miss those rides. Fortunately it seems, that most of the early arrivals happen at the end of each trip and therefor don't affect so many people.

```{r tram_2, echo=FALSE}
plot_tram_arrival("2")
```

#### Tram number 3
<br>
From 11:30 - 13:30 almost all the trams leaving from Tondi to Kadriorg (10 in total) are late ~10 minutes. The delays are caused in the sections before Tallinn-Väike and Vineeri stop. The accumulated late arrivals carry on with the trams up to 16:00. 3 trams leaving Kadriorg are late around 18:00 (as were 4 trams number 1), but all in all the rush hour tram rides are quite on time.

```{r tram_3, echo=FALSE}
plot_tram_arrival("3")
```

#### Tram number 4
<br>
The patter of delays for tram number 4 are quite similar to tram number 3. As the two trams share the same route from Tondi to Hobujaama stop, major delays for tram number 4 are also happening before Vineeri stop from 12:00 - 13:00. Actually this place for delays is quite weird, as the tram line is separated from driveway over there. Possibly there was a car crash, that affected all the trams between Tallinn-Väike and Vineeri stop. Rush hour arrivals around 17:00 are on time.

```{r tram_4, echo=FALSE}
plot_tram_arrival("4")
```


### Further analysis

To make some more solid conclusions on the tram delays, it is definitely necessary to collect data for more than one day and to use a longer period of time to choose the data collection days from. It would also be very interesting to do a similar analysis on buses delays, as they are probably much more affected by the overall traffic (separated bus lines are only in city center and on some major roads).

<br>